V4 Neural Network Model for Shape-based Feature Extraction and Object Discrimination

### 2015-09-05 article.tex
finish experiment part.

\documentclass[9pt,twocolumn]{article}

@@

Understanding the content of images has always been a difficult task in image analysis due to the well known semantic gap \cite{smeulders2000}.
A high degree of abstraction is required to transform low-level pixel data into high-level knowledge of the images.
In an attempt to bridge this semantic gap, various methods have been proposed to extract features from the pixel representation of images.
However, humans and primates outperform any existing computer vision system with respect to almost any measure.
Though the neural mechanism of vision has not been fully revealed,
recent research on neuroscience do provide many inspiring discoveries which would help development of computer vision algorithms.
In this paper, we demonstrate a novel model of visual area V4 and its application to shape-based feature extraction and object discrimination.
Experiment results show that the proposed model exhibits similar shape selectivity to that of real V4 neurons
and it outperforms many existing methods in computer vision tasks including feature matching and image classification.

@@

According to the hierarchy of the ventral visual pathway, area V4 receives input from the lower levels including area V1 and V2.
Neurons in V1 and V2 respond to local orientations.
They fall into two categories, simple cells and complex cells.
A simple cell is activated when an edge stimulus is presented in its receptive field with appropriate orientation and position.
Complex cells have larger receptive fields.
A stimulus is effective wherever it is placed in the complex receptive field, provided that the orientation is appropriate \cite{hubel1962}.

In our model, the layer of simple cells covers the whole dimension of input images.
For each location, several simple cells with different preferred orientations detect edges of different orientations.
Simple cells are modeled as Gabor functions,
\begin{equation}\label{equ:gabor}
g_{\theta}(x,y;\lambda,\sigma_s)
=\exp \left(-\frac{x'^2+y'^2}{2\sigma_s^2}\right)
\cos \left(2\pi\frac{x'}{\lambda}+\psi\right),
\end{equation}
where $x'=x\cos\theta+y\sin\theta$, $y'=-x\sin\theta+y\cos\theta$.
In the equation, $\lambda$ is the wavelength of the sinusoidal factor, 
$\psi$ is the phase offset, 
$\theta$ represents the preferred orientation and 
$\sigma_s$ approximates the radius of the receptive fields.

\begin{figure}
\centering
\subfloat[Phase offset $\psi=\pi/2$]{
  \includegraphics[width=0.4\linewidth]{images/fig4a.jpg}
  \label{fig:4a}}
\subfloat[Phase offset $\psi=0$]{
  \includegraphics[width=0.4\linewidth]{images/fig4b.jpg}
  \label{fig:4b}}
\caption{Gabor functions with different phases.}
\label{fig:4}
\end{figure}

It is notable that the phase offset $\psi$ affects the pattern of the receptive fields.
\figurename~\ref{fig:4}~shows two Gabor functions with different phase offsets.
The sinusoidal factors of the two Gabor functions are odd function and even function respectively.
The odd Gabor function resembles a gradient operator while the even Gabor function resembles a Laplacian operator.
In a computational point of view, the Laplacian operator calculates a higher order derivative than the gradient operator.
The gradient operator suppresses the smooth area in images.
Moreover, the Laplacian operator is able to suppress area that varies smoothly, which is also neglected by human eyes generally.
Therefore, in our model, the even Gabor function with the phase offset $\psi=0$ is employed.

The layer of simple cells operates on raw image input. 
The output of simple cells with preferred orientation $\theta$ and scale $\sigma_s$ is the following convolution passed through a transfer function $\phi$,
\begin{equation}\label{equ:gabor}
S_{\theta,\sigma_s}(x,y)=\phi(I\otimes g_{\theta,\sigma_s}),
\end{equation}
where $I$ is an image and 
\begin{equation}\label{equ:gabor}
\phi(u)=\left\{\begin{array}{ll}
u & \text{if } u>0,\\
0 & \text{otherwise.}
\end{array}\right.
\end{equation}

@@

\figurename~\ref{fig:5} shows four image patches.
For each image patch, the output values of complex cells with different preferred orientations (from $0^\circ$ to $180^\circ$) are plotted in a bar chart.
The charts show the distribution of complex cell activities.
The entropy calculated accordingly indicates that patches composed of simple structures have non-uniform distributions and thus low entropy values.
Therefore, we filter out regions with high entropy.

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{images/fig5.jpg}} 
\caption{Entropy of image patches.
Bar charts in the second row show the output values of complex cells
with different preferred orientations (from $0^\circ$ to $180^\circ$).}
\label{fig:5}
\end{figure}

In addition to entropy, local competition also plays a role in attentional selection \cite{desimone1995}.
With limited neural resources, only strong and competitive neuronal signals get transmitted and processed.
In our model, the saliency layer finds local maximums of complex cell output
and filters out those points with high entropy values or low activities (or output values).
The algorithm is listed in Algorithm \ref{alg:1}.
In our experiment, the saliency filters take a threshold $t_E=2.8$ for the entropy
and a threshold $t_A=0.4$ for complex cell output value (or complex cell activity).
These values are roughly the median values of natural images.

@@

@@

\section{Shape-based feature extraction}\label{sec:4}

In this section, we demonstrate a series of experiments.
In the experiments, our proposed model extracts local shape features from images.
Compared with V4 neurons, it shows similar selectivity for different features.

\subsection{Perceptron over Complex Cells}

We have reviewed the shape selectivity of V4 neurons in section \ref{sec:2}.
In the following experiment, we examined that the output of complex cells provides sufficient information
for the emergence of neuronal response pattern of V4.
The experiment involved only the first two layers (i.e., simple cells and complex cells) in our model.
A perceptron was configured to take complex cells as input.
It was then trained to distinguish between two shapes shown in \figurename~\ref{fig:7a}.
The shapes were also used to examine the selectivity of V4 in \cite{pasupathy2001}.
The difference between the two shapes is that one has a sharp projection towards the top right.

\begin{figure}
\centering
\subfloat[Response map]{\includegraphics[width=0.48\linewidth]{images/fig-7a.jpg}\label{fig:7a}}\hfil
\subfloat[Weight matrices]{\includegraphics[width=0.48\linewidth]{images/fig-7b.jpg}\label{fig:7b}}
\caption{Response map and weight matrices of a perceptron that distinguishes two shapes.
(a) Responses of the perceptron over 18 samples.
The perceptron prefers the shape in the left half of the samples.
It is insensitive to stimulus position.
(b) The input weight of the perceptron. 
Each block shows the weight of connections from complex cells of a certain orientation.
Complex cells of 18 different orientations provide input for this perceptron.}
\label{fig:7}
\end{figure}

Since V4 neurons show a certain degree of invariance in their selectivity across changes in stimulus position,
we moved the shapes randomly within the receptive field of the perceptron 
to generate samples for training and testing (\figurename~\ref{fig:7a} shows several samples).
The samples were then used as input of the layers of simple cells and complex cells.
The output of complex cells were passed to the perceptron.
We had complex cells with different preferred orientations 
(from $0^\circ$ to $170^\circ$ in steps of $10^\circ$) and over different positions in the receptive field.
Therefore, the input of the perceptron consisted of 18 matrices, 
each corresponding to the output of complex cells with a certain orientation.
The input weight of the perceptron was thus also 18 matrices.

The trained perceptron exhibited a strong bias towards the shape with convex curvature towards the top right.
It also showed a significant degree of invariance to the stimulus position.
It is obvious that the selectivity is not formed from certain excitatory sub-regions
or inhibitory sub-regions of the receptive field which was found of simple cells \cite{hubel1962,pasupathy2001}.
The selectivity of the perceptron tallies with the selectivity of actual V4 neurons.
The response map is shown in \figurename~\ref{fig:7a}.
Darker background colors indicate stronger responses.
The input weight of this perceptron is shown in \figurename~\ref{fig:7b}.
Each block shows the weight from the complex cells of a certain preferred orientation.
Red color denotes positive weight while blue color denotes negative weight.

@@

\subsection{Shape Selectivity}

In the following experiment, we trained our model to learn shapes.
The training process involved the layers of simple cells, complex cells and RBM encoders.
The training samples were collected from \cite{gallant1996,pasupathy1999,pasupathy2001}, 
including 4 categories of stimuli, i.e.,
sinusoidal gratings, non-Cartesian gratings, segmented curves, and closed shapes.
The simple cells in our model took these sample images as input.
The images were processed by simple cells and complex cells.
Since the sample images were fitted into V4 receptive fields in neurobiological experiments,
we adjusted the scale of the images to fit the size of a single V4 receptive field in our model too.
Therefore the saliency filters were not necessary in this experiment.
The output of complex cells were passed directly on to an RBM encoder.

We used an RBM with 256 hidden units (RBM output neurons) in this experiment.
In order to assess the selectivity of these units,
we assigned each unit a selectivity index over a category of stimuli.
The selectivity index is defined as the ratio of the maximal response to the average response 
over the stimuli of a certain category.
Given an RBM output unit $h$ and its output value $h_i$ over a category of $N$ stimuli for $i=1,2,\dots,N$.
The selectivity index $\mathcal{S}$ of $h$ is given by
\begin{equation}
\mathcal{S}(h)=\frac{\max\{h_i|i=1,2,\dots,N\}}{\sum_{i=1}^N h_i/N}.
\end{equation}
In the following statistics, we assumed that a neuron has significant selectivity over a category of stimuli
if the selectivity index is greater than 3.5.
Among the 256 units, 171 units exhibited significant selectivity for different stimuli.
102 units showed selectivity for segmented curves (\figurename~\ref{fig:8}).
55 units showed strong bias towards non-Cartesian gratings (\figurename~\ref{fig:9})
while 50 units showed selectivity for classical sinusoidal gratings.
68 units exhibited selectivity for closed shapes (\figurename~\ref{fig:10}).
The shape selectivity of our model is listed in \tablename~\ref{tab:2}.
The categories of selectivity are not mutually disjoint.
A single unit may possess more than one kind of selectivity simultaneously.

\begin{table}[h]
\caption{Shape selectivity of our model}
\centering
\begin{tabular}{rrr}
\toprule
Selectivity & Number of Units & Percentage \\
\midrule
Sinusoidal gratings & 50 & 19.5\% \\
Non-Cartesian gratings & 55 & 21.5\% \\
Segmented curves & 102 & 39.8\% \\
Closed shapes & 68 & 26.6\% \\
\bottomrule
\end{tabular}
\label{tab:2}
\end{table}

\figurename~\ref{fig:8a} shows the sample patterns of segmented curves.
\figurename~\ref{fig:8b} to \figurename~\ref{fig:8d} show the response maps of 3 RBM output neurons.
Darker background indicates stronger response.
The selectivity indexes is shown at the top.
It is shown that the three neurons exhibited strong bias towards curves rather than straight lines.
The neurons were tuned for the orientation of the projection of the segmented curves.
This is compliant with the selectivity of V4 neurons described in \cite{pasupathy1999}.

\figurename~\ref{fig:9} shows the response maps of two RBM output neurons
which are selective for non-Cartesian gratings.
Both of the two neurons were not sensitive to sinusoidal gratings.
The one in \figurename~\ref{fig:9a} was selective for a pair of circular sectors.
The one in \figurename~\ref{fig:9b} preferred non-Cartesian gratings of helix shapes.

\figurename~\ref{fig:10} shows the sample patterns of closed shapes and the response map of two neurons.
The two neurons were selectively tuned for two kinds of shapes.
They were sensitive to the angular position of the shapes.

\begin{figure}
\centering
\subfloat[Patterns]{\includegraphics[width=0.24\linewidth]{images/fig-8a.jpg}\label{fig:8a}}\hfil
\subfloat[Response map]{\includegraphics[width=0.24\linewidth]{images/fig-8b.jpg}\label{fig:8b}}\hfil
\subfloat[Response map]{\includegraphics[width=0.24\linewidth]{images/fig-8c.jpg}\label{fig:8c}}\hfil
\subfloat[Response map]{\includegraphics[width=0.24\linewidth]{images/fig-8d.jpg}\label{fig:8d}}
\caption{RBM output neurons responding to convex curves towards certain directions.
(a) Sample patterns of segmented curves.
(b) Response map of a neuron tuned for curves projecting upwards.
(c) Response map of a neuron tuned for curves projecting downwards.
(d) Response map of a neuron tuned for curves projecting towards the top left.}
\label{fig:8}
\end{figure}

\begin{figure}
\centering
\subfloat[Response map of a neuron selective for a pair of circular sectors (selectivity index = 3.92).]%
{\includegraphics[width=0.9\linewidth]{images/fig-9a.jpg}\label{fig:9a}}\\
\subfloat[Response map of a neuron selective for helix shapes (Selectivity index = 6.37).]%
{\includegraphics[width=0.9\linewidth]{images/fig-9b.jpg}\label{fig:9b}}
\caption{Response maps of RBM output neurons tuned for non-Cartesian gratings.}
\label{fig:9}
\end{figure}

\begin{figure}
\centering
\subfloat[Patterns]{\includegraphics[width=0.31\linewidth]{images/fig-10a.jpg}\label{fig:10a}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig-10b.jpg}\label{fig:10b}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig-10c.jpg}\label{fig:10c}}
\caption{RBM output neurons responding to closed shapes.}
\label{fig:10}
\end{figure}

This experiment demonstrates that our model can learn to distinguish the stimuli
which V4 neurons are selectively tuned for.

\subsection{Feature detection}

In the above experiments, it has been demonstrated that a unit of the RBM hidden layer 
can encode the shape feature in the receptive field of a single V4 computation unit.
When we deal with larger images, especially natural images,
a full scan over the whole image is not computationally economic.
It also complicates succeeding processing for high-level tasks such as object recognition and scene understanding.
We have to find some regions of interest 
and focus the computation units over a limited number of interesting regions.
The visual neural system takes a similar approach.
As we have reviewed in previous sections, V4 is closely related with selective visual attention \cite{roe2012}.
Research shows that V4 receptive field shrinks and shifts towards saccade target \cite{hamker2006}.
We simulated such visual attention with the saliency filters in our model.
In the following experiment,
we evaluated the saliency filters with feature detection experiment.

\begin{figure}
\centering
\subfloat[Car image]{\includegraphics[width=0.32\linewidth]{images/fig-11a.jpg}\label{fig:11a}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig-11c.jpg}\label{fig:11b}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig-11e.jpg}\label{fig:11c}}\\
\subfloat[Face image]{\includegraphics[width=0.32\linewidth]{images/fig-11b.jpg}\label{fig:11d}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig-11d.jpg}\label{fig:11e}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig-11f.jpg}\label{fig:11f}}\\
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig-11g.jpg}\label{fig:11g}}\hfil
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig-11h.jpg}\label{fig:11h}}
\caption{Feature detection. (a) and (d) show the original images.
(b) and (e) are the complex cell activity. 
Red color indicates strong activity and blue color indicates inhibition.
(c) and (f) are the entropy map.
Red color indicates high entropy value and blue color indicates low entropy value.
(g) and (h) are the selected salient points
at the scale of 4, 8, and 16 in terms of simple cell radius.}
\label{fig:11}
\end{figure}

\figurename~\ref{fig:11} shows the process of feature detection with the saliency filters in our model.
\figurename~\ref{fig:11b} and \figurename~\ref{fig:11d} show the activity map of complex cells.
The activities are high in the region of the trees behind the car (\figurename~\ref{fig:11a})
and the handwriting on the white board (\figurename~\ref{fig:11c}).
Saliency filters filter out these regions because these regions have a comparatively high entropy.
These areas consist of a large amount of disordered edges
which result in a near uniform distribution of complex cell activity over preferred orientations
and thus high entropy values.
The saliency filters can find salient points at each given scale.
\figurename~\ref{fig:11g} and \figurename~\ref{fig:11h} show the salient points
at the scale of 4, 8, and 16 in terms of the simple cell radius
(The size of the other units can be inferred with the relationship
shown in Table \ref{tab:1}).

\section{Object discrimination}
\label{sec:5}

In this section, we demonstrate the performance of the features 
generated by our model in feature matching and object recognition tasks.

We do not use typical criteria for evaluating local descriptors \cite{mikolajczyk2005}
for the reason that the typical criteria focus on
matching the same physical point in different viewing conditions,
at which humans or other intelligent species
are not significantly better than computer systems.
The human vision excels in image understanding tasks
such as recognizing a new instance of some category
with only few prototypes for learning.
Our model conforms to the neural mechanism of vision and
thus exhibits similar learning ability in the experiments.

@@

Traditional local features excel in matching features
corresponding to the same physical point on an object
seen from different viewpoints \cite{mikolajczyk2005}.
These features are usually very selective for a target object
but lack invariance with respect to different objects from a target category.
However, the generalization ability within object categories
is important for image understanding tasks such as object recognition.

We provide a qualitative evaluation of our features
by matching features between different images.
The SIFT features are used for comparison.
The VLFeat open-source library \cite{vedaldi2008}
were utilized for the SIFT implementation 
and the feature matching algorithm.

\begin{figure}
\centering
  \subfloat[Matching between the same face]{
    \includegraphics[width=0.8\linewidth]{images/fig-12a.jpg}
    \label{fig:12a}}\\
  \subfloat[Matching between different faces]{
    \includegraphics[width=0.8\linewidth]{images/fig-12b.jpg}
    \label{fig:12b}}
\caption{Matching face images.}
\label{fig:12}
\end{figure}

\figurename~\ref{fig:12} shows matchings of features in face images.
As shown in \figurename~\ref{fig:12a},
SIFT features match perfectly between face images of the same person.
The neural features generated by our model achieve competitive distinctiveness.
However, \figurename~\ref{fig:12b} shows that
matching of SIFT features between face images of different people
is quite disordered.
SIFT features excel in matching highly distinctive features under image transformations 
but lack the generalization ability 
to capture variations in objects appearance of the same category.
By contrast, the neural features show robust performance 
in matching features from different faces.
The result suggests that our model exhibits 
both competitive discrimination ability
and better generalization ability compared with SIFT features.

\begin{figure}
\centering
  \subfloat[Our match]{
    \includegraphics[width=0.8\linewidth]{images/fig-13a.jpg}
    \label{fig:13a}}\\
  \subfloat[SIFT match]{
    \includegraphics[width=0.8\linewidth]{images/fig-13b.jpg}
    \label{fig:13b}}
\caption{Comparison of feature matching between van images.}
\label{fig:13}
\end{figure}

Another comparison in \figurename~\ref{fig:13} demonstrates
more clearly the advantage of our model compared with SIFT feature.
\figurename~\ref{fig:13a} shows that our model produced correct matches
between the model van in the left and the two real vans in the right. 
However, SIFT feature failed to produce correct matches (\figurename~\ref{fig:13b}). 
SIFT feature utilizes the direction of image gradient. 
It fails to capture the shape feature of an object when
the object's color changes dramatically because the direction of
image gradient changes with the color. 
Our model focuses on the shape feature and thus achieves a stable matching despite
the change in color.

\subsection{Object recognition layer}

>>>>

\documentclass[twocolumn]{article}

@@

Understanding the content of images has always been a difficult task in image analysis due to the well known semantic gap \cite{smeulders2000}.
A high degree of abstraction is required to transform low-level pixel data into high-level knowledge of the images.
In an attempt to bridge this semantic gap, various methods have been proposed to extract features from the pixel representation of images.
However, humans and primates outperform any existing computer vision system with respect to almost any measure.
Though the neural mechanism of vision has not been fully revealed,
recent research on neuroscience do provide many inspiring discoveries which would help development of computer vision algorithms.
In this paper, we demonstrate a novel model of visual area V4 and its application to shape-based feature extraction and object discrimination.
Experiment results show that the proposed model exhibits similar shape selectivity to that of real V4 neurons
and it outperforms many existing methods in computer vision tasks including feature matching and object recognition.

@@

According to the hierarchy of the visual cortex, area V4 receives input from the lower levels including area V1 and V2.
Neurons in V1 and V2 respond to local orientations.
They fall into two categories, simple cells and complex cells.
A simple cell is activated when an edge stimulus is presented in its receptive field with appropriate orientation and position.
Complex cells have larger receptive fields.
A stimulus is effective wherever it is placed in the complex receptive field, provided that the orientation is appropriate \cite{hubel1962}.

In our model, simple cells are modeled as Gabor functions,
\begin{equation}\label{equ:gabor}
g_{\theta}(x,y;\lambda,\sigma_s)
=\exp \left(-\frac{x'^2+y'^2}{2\sigma_s^2}\right)
\cos\frac{2\pi x'}{\lambda},
\end{equation}
where $x'=x\cos\theta+y\sin\theta$, $y'=-x\sin\theta+y\cos\theta$.
In the equation, $\lambda$ is the wavelength of the sinusoidal factor, 
$\theta$ represents the preferred orientation,
and $\sigma_s$ approximates the radius of the receptive fields.

Simple cells operates on raw image input. 
The output of simple cells with preferred orientation $\theta$ and scale $\sigma_s$ is the following convolution passed through a transfer function $\phi$,
\begin{equation}\label{equ:gabor}
S_{\theta,\sigma_s}(x,y)=\phi(I\otimes g_{\theta,\sigma_s}),
\end{equation}
where $I$ is an image and 
\begin{equation}\label{equ:gabor}
\phi(u)=\left\{\begin{array}{ll}
u & \text{if } u>0,\\
0 & \text{otherwise.}
\end{array}\right.
\end{equation}

@@

\figurename~\ref{fig:4} shows four image patches.
For each image patch, the output values of complex cells with different preferred orientations (from $0^\circ$ to $180^\circ$) are plotted in a bar chart.
The charts show the distribution of complex cell activities.
The entropy calculated accordingly indicates that patches composed of geometrically ordered structures have non-uniform distributions and thus low entropy values.
Therefore, we filter out regions with high entropy.

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{images/fig4.jpg}} 
\caption{Entropy of image patches.
Bar charts in the second row show the output values of complex cells
with different preferred orientations (from $0^\circ$ to $180^\circ$).}
\label{fig:4}
\end{figure}

In addition to entropy, local competition also plays a role in attentional selection \cite{desimone1995}.
In our model, the saliency layer finds local maximums of complex cell output and filters out those points with high entropy values or low activities.
The algorithm is listed in Algorithm \ref{alg:1}.
In our experiment, the saliency filters take a threshold $t_E=2.8$ for the entropy
and a threshold $t_A=0.4$ for complex cell output value (or complex cell activity).
These values are roughly the median values of natural images.

@@

\figurename~\ref{fig:5} shows the process of feature detection.
Saliency filters successfully filter out the trees and handwriting at the background of the images.
These regions have little shape features but the activity map of complex cells contains high values caused by a large amount of disordered edges. 
Disordered edges result in high entropy values because the distribution of complex cell activity over different orientations is near uniform.
Saliency filters filter out these regions by applying a threshold to the entropy map.

\begin{figure}
\centering
\subfloat[Car image]{\includegraphics[width=0.32\linewidth]{images/fig5a.jpg}\label{fig:5a}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig5c.jpg}\label{fig:5b}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig5e.jpg}\label{fig:5c}}\\
\subfloat[Face image]{\includegraphics[width=0.32\linewidth]{images/fig5b.jpg}\label{fig:5d}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig5d.jpg}\label{fig:5e}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig5f.jpg}\label{fig:5f}}\\
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig5g.jpg}\label{fig:5g}}\hfil
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig5h.jpg}\label{fig:5h}}
\caption{Feature detection. (a)(d) Original images.
(b)(e) Complex cell activity map. 
(c)(f) Entropy map.
In the maps, red color indicates high value and blue color indicates low value.
(g)(h) Salient points at the scale of 4, 8, and 16 in terms of simple cell radius.}
\label{fig:5}
\end{figure}

@@

\section{Experiments}\label{sec:4}

In this section, we demonstrate several experiments.
The first two experiments show that the shape selectivity of our model conforms to that of real V4 neurons.
The following experiments evaluate our model in feature matching and object recognition tasks.

\subsection{Perceptron over complex cells}

With this experiment, we demonstrate that artificial neurons which take input directly from complex cells can imitate neuronal response pattern of V4.
Therefore, it is feasible that we use a single layer of RBM to extract V4 features from complex cells.

In the experiment, a perceptron was configured to take complex cells as input.
It was then trained to distinguish between two shapes shown in \figurename~\ref{fig:7a}.
The shapes were also used to examine the selectivity of V4 in \cite{pasupathy2001}.
The difference between the two shapes is that one has a sharp corner towards the top right.

\begin{figure}
\centering
\subfloat[Response map]{\includegraphics[width=0.48\linewidth]{images/fig7a.jpg}\label{fig:7a}}\hfil
\subfloat[Weight matrices]{\includegraphics[width=0.48\linewidth]{images/fig7b.jpg}\label{fig:7b}}
\caption{Response map and weight matrices of a perceptron that distinguishes two shapes.
(a) Responses of the perceptron over 18 samples.
The perceptron prefers the shape in the left half of the samples.
It is insensitive to stimulus position.
(b) The input weight of the perceptron. 
Each block shows the weight of connections from complex cells of a certain orientation.
Complex cells of 18 different orientations provide input for this perceptron.}
\label{fig:7}
\end{figure}

Since V4 neurons show a certain degree of invariance in stimulus position,
we moved the shapes randomly within the receptive field of the perceptron 
to generate samples for training and testing (\figurename~\ref{fig:7a} shows several samples).
The samples were then used as input of the layers of simple cells and complex cells.
The output of complex cells were passed to the perceptron.
We had complex cells with 18 different preferred orientations (from $0^\circ$ to $170^\circ$ in steps of $10^\circ$) 
and over different positions in the V4 receptive field.
Therefore, the input of the perceptron consisted of 18 matrices, 
each corresponding to the output of complex cells with a certain orientation.
The input weight of the perceptron was thus also 18 matrices.

The trained perceptron exhibited a strong bias towards the shape with a sharp corner towards the top right.
It also showed a significant degree of invariance to the stimulus position.
It is obvious that this kind of selectivity is not formed from certain excitatory sub-regions
or inhibitory sub-regions of the receptive field which was found of simple cells \cite{hubel1962,pasupathy2001}.
The selectivity of the perceptron tallies with the selectivity of actual V4 neurons.
The response map is shown in \figurename~\ref{fig:7a}.
Darker background colors indicate stronger responses.
The input weight of this perceptron is shown in \figurename~\ref{fig:7b}.
Each block shows the weight from the complex cells of a certain preferred orientation.
Red color denotes positive weight while blue color denotes negative weight.

@@

\subsection{Shape selectivity}

In the following experiment, we trained our model to learn shapes.
After training, our model can distinguish the stimuli which V4 neurons are selectively tuned for.
The training samples were collected from \cite{gallant1996,pasupathy1999,pasupathy2001}, 
including 4 categories of stimuli, i.e., sinusoidal gratings, non-Cartesian gratings, segmented curves, and closed shapes.
The training process involved the layers of simple cells, complex cells and RBM encoders.
Since each sample image was fitted into a single V4 receptive field, the saliency filters were not involved in this experiment.

We used an RBM with 256 hidden units in this experiment.
In order to assess the selectivity of these units, we assigned each unit a selectivity index over a category of stimuli.
The selectivity index is defined as the ratio of the maximal response to the average response over the stimuli of a certain category.
Given an RBM output unit $h$ and its output value $h_i$ over the $i$-th stimulus in a category of $N$ stimuli.
The selectivity index $\mathcal{S}$ of $h$ is given by
\begin{equation}
\mathcal{S}(h)=\frac{\max\{h_i|i=1,2,\dots,N\}}{\sum_{i=1}^N h_i/N}.
\end{equation}
In the following statistics, we assumed that a unit has significant selectivity over a category of stimuli
if its selectivity index is greater than the average index.
Among the 256 units, 171 units exhibited significant selectivity for different stimuli.
102 units showed selectivity for segmented curves (\figurename~\ref{fig:8}).
55 units showed strong bias towards non-Cartesian gratings (\figurename~\ref{fig:9})
while 50 units showed selectivity for classical sinusoidal gratings.
68 units exhibited selectivity for closed shapes (\figurename~\ref{fig:10}).
The shape selectivity of our model is listed in \tablename~\ref{tab:1}.
The categories of selectivity are not mutually disjoint.
A single unit may possess more than one kind of selectivity simultaneously.

\begin{table}[h]
\caption{Shape selectivity of our model}
\centering
\begin{tabular}{rrr}
\toprule
Selectivity & \#Units & Percentage \\
\midrule
Sinusoidal gratings & 50 & 19.5\% \\
Non-Cartesian gratings & 55 & 21.5\% \\
Segmented curves & 102 & 39.8\% \\
Closed shapes & 68 & 26.6\% \\
\bottomrule
\end{tabular}
\label{tab:1}
\end{table}

\figurename~\ref{fig:8} show three response maps to segmented curves.
Darker background indicates stronger response.
It is shown that the three units exhibited strong bias towards curves rather than straight lines.
They were all tuned for the orientation of the projections.
This is compliant with the selectivity of V4 neurons described in \cite{pasupathy1999}.
\figurename~\ref{fig:9} shows the response maps of two units which are selective for non-Cartesian gratings.
Both of the two units were not sensitive to sinusoidal gratings.
\figurename~\ref{fig:10} shows the sample patterns of closed shapes and the response map of two units selectively tuned for two kinds of shapes.
They were sensitive to the angular position of the shapes \cite{pasupathy2001}.

\begin{figure}
\centering
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8a.jpg}\label{fig:8a}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8b.jpg}\label{fig:8b}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8c.jpg}\label{fig:8c}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8d.jpg}\label{fig:8d}}
\caption{RBM output units responding to convex curves towards certain directions.
(a) Sample patterns of segmented curves.
(b-d) Selectivity indexes and response maps of units tuned for upward, downward, and top-left projections. }
\label{fig:8}
\end{figure}

\begin{figure}
\centering
\subfloat[Response map of a unit selective for a pair of circular sectors (selectivity index = 3.92).]%
{\includegraphics[width=0.9\linewidth]{images/fig9a.jpg}\label{fig:9a}}\\
\subfloat[Response map of a unit selective for helix shapes (Selectivity index = 6.37).]%
{\includegraphics[width=0.9\linewidth]{images/fig9b.jpg}\label{fig:9b}}
\caption{Response maps of RBM units tuned for non-Cartesian gratings.}
\label{fig:9}
\end{figure}

\begin{figure}
\centering
\subfloat[Patterns]{\includegraphics[width=0.31\linewidth]{images/fig10a.jpg}\label{fig:10a}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig10b.jpg}\label{fig:10b}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig10c.jpg}\label{fig:10c}}
\caption{RBM units responding to closed shapes.}
\label{fig:10}
\end{figure}

@@

In this section, we demonstrate the performance of the features generated by our model in feature matching tasks.
We do not use typical criteria for evaluating local descriptors \cite{mikolajczyk2005}
for the reason that the typical criteria focus on matching the same physical point in different viewing conditions,
at which humans or other intelligent species are not significantly better than computer systems.
The human vision excels in image understanding tasks such as recognizing a new instance from some category.
It requires generalization ability of the features.

We first provide a qualitative evaluation.
The SIFT features are used for comparison.
The VLFeat open-source library \cite{vedaldi2010} were utilized for the SIFT implementation and the feature matching algorithm.

\begin{figure}
\centering
  \subfloat[Matching between the same face]{
    \includegraphics[width=0.8\linewidth]{images/fig11a.jpg}
    \label{fig:11a}}\\
  \subfloat[Matching between different faces]{
    \includegraphics[width=0.8\linewidth]{images/fig11b.jpg}
    \label{fig:11b}}
\caption{Matching face images.}
\label{fig:11}
\end{figure}

\figurename~\ref{fig:11} shows matchings of features in face images.
As shown in \figurename~\ref{fig:11a}, both SIFT features and neural features match perfectly between face images of the same person.
However, \figurename~\ref{fig:11b} shows that matching of SIFT features between face images of different people is quite disordered.
SIFT features excel in matching highly distinctive features under image transformations
but lack the generalization ability to capture variations in objects appearance of the same category.
By contrast, the neural features show robust performance in matching features from different faces.
The result suggests that our model exhibits both competitive discrimination ability and better generalization ability compared with SIFT features.

\begin{figure}
\centering
  \subfloat[Our match]{
    \includegraphics[width=0.8\linewidth]{images/fig12a.jpg}
    \label{fig:12a}}\\
  \subfloat[SIFT match]{
    \includegraphics[width=0.8\linewidth]{images/fig12b.jpg}
    \label{fig:12b}}
\caption{Comparison of feature matching between van images.}
\label{fig:12}
\end{figure}

Another comparison in \figurename~\ref{fig:12} demonstrates
more clearly the advantage of our model compared with SIFT feature.
\figurename~\ref{fig:12a} shows that our model produced correct matches
between the model van in the left and the two real vans in the right. 
However, SIFT feature failed to produce correct matches (\figurename~\ref{fig:12b}). 
SIFT feature utilizes the direction of image gradient. 
It fails to capture the shape feature of an object when the object's color changes dramatically because the direction of image gradient changes with the color. 
Our model focuses on the shape feature and thus achieves a stable matching despite the change in color.

For quantitative evaluation, we use the ``Face-easy'' category from the Caltech 101 dataset \cite{fei2007}.
This category consists of frontal face photos of many different people.
They are all cropped to the same size and the same face location.
The images are randomly paired for feature matching.
A correct match between facial features should be corresponding regions in the two images.
We use precision-recall curves to compare our neural features with the SIFT features.
\begin{equation}
\text{Precision} = \frac{\text{\#correct matches}}{\text{\#correct matches} + \text{\#false matches}}.
\end{equation}
\begin{equation}
\text{Recall} = \frac{\text{\#correct matches}}{\text{\#all features}}.
\end{equation}
The result is shown in \figurename~\ref{fig:13}.
Extensive evaluation on the generalization ability of our features is in the object recognition experiment in the next subsection.

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{images/fig13.jpg}} 
\caption{Matching facial features in Caltech-101 ``Face-easy'' category.}
\label{fig:13}
\end{figure}
